Recurrent event analysis in the presence of real-time high frequency data via random subsampling

### 2020-08-12 write-up/fda-recurrent.tex
updated

In this section, we extend our model to the case of multiple functional regressors.  That is, suppose $L$ health processes, i.e., $\bfx_i = \{ \bfx_i (t) = (x_{i,1} (t), \ldots, x_{i,L} (t)) \}_{0 < s < \tau_i}$, for each participant is measured at a dense grid of time points. In the suicidal ideation case study, for example, accelerometer is measured at a rate of 32Hz while electrodermal activity (EDA) is measured at a rate of 4Hz.  A multivariate extension of our model~\eqref{eq:hazard} is given by
\begin{equation}
\label{eq:multihazardlinear}
h_i \left( t \given  H_{i,t}^{NX} ; \theta \right) =
h_0 (t; \gamma) \exp \left( g_t \left( H_{i,t}^{N} \right)^{\prime} \alpha
  + \int_{t-\Delta}^t x_{i,1} (s) \beta_1(s) ds + \cdots + \int_{t-\Delta}^t x_{i,L} (s) \beta_2(s) ds  \right)
\end{equation}
The approach given in Section~\ref{eq:logistication} extends naturally to the multivariate functional setting. For each functional regressor, we estimate the pooled sample covariance $\Sigma_{y,l}$ for $y \in \{0,1\}$ and $l=1,\ldots,L$ as in Section~\ref{sec:margcov}.  Let $\sum_{k=1}^\infty \hat \lambda^{(y)}_{k,l} \hat \psi^{(y)}_{k,l} (s) \hat \psi_{k,l}(t)$ be the spectral decomposition of $\hat \Sigma_{y,l}$.  Then $x_{i,j}(t)$ is approximated using a truncated Karhunen-Loeve

, the coefficient is expressed using separate marginal covaraiance
$$
\int_0^\Delta x_{il} \beta(t) =
$$

Note that we express each coefficient function in the same spline basis; however, different basis expansions are used for each health process as defined by the marginal covariance calculation in Section~\ref{sec:margcov}.

@@

Here we present an extension of~\cite{Petrovich2018} to our setting by leveraging Lemma~\ref{lemma:logistic} and the marginal covariance estimation procedure to construct a multiple imputation procedure. Let~$\bfx_{i} (t)$ denote incomplete sensor data at time~$t$ (i.e., at times~$\{ s_{i,r} \}_{r=1}^{k_{it}} \}$ in~$[0,\Delta]$. Then
\begin{align}
\label{eq:impute}
\E [ X_i (s,t) \given Y_i(t) = y, \bfx_i (t) ]
  &= \mu_y (s,t) + {\bf a}_{i,t}^\top (s) {\bf B}_{i,t} (\bfx_i (t) -
    \mathbold{\mu}_i (t) ) \\
\text{Cov} \left( X_i (s,t), X_i (s^\prime, t) \given Y_i (t) = y,
  \bfx_i (t) \right)
  &= \Sigma_{t} (s, s^\prime) -
    {\bf a}_{i,t} (s)^\top {\bf B}_{i,t} {\bf a}_{i,t} (s^\prime)
\end{align}
where we have
\[
{\bf a}_{i,t} (s)^\top = \left( \begin{array}{c} \Sigma_t (s_{i,1}, s) \\ \vdots \\
                       \Sigma_t (s_{i,k_{it}}, s) \end{array} \right); \quad
{\bf B}_{i,t}^{-1} = \left(
  \begin{array}{c c c}
    \Sigma_t(s_{i,1}, s_{i,1}) & \Sigma_t(s_{i,1}, s_{i,2}) & \cdots \\
    \Sigma_t(s_{i,1}, s_{i,2}) & \ddots & \vdots \\
    \vdots & \cdots & \Sigma_t (s_{i,k_{it}}, s_{i,k_{it}})
  \end{array} \right),
\]
$\mathbold{\mu}_i (t) = \E [ \bfx_i (t) \given Y_i (t) = y] = \{
\mu_y (s_{j}, t) \}_{j=1}^r$ and~$\mu_y (s,t)$ is the mean of~$X(t,s)$
from group~$y$, and~$\Sigma_t ( s^\prime, s)$ is the covariance
between~$X(s^\prime,t)$ and $X(s,t)$ for $s^\prime, s \in [0, \Delta]$
and~$t \in \mathbb{R}_+$.
These covariances can be estimated using the methods considered by
\cite{ChenMuller2012}.
% Our imputation procedure instead uses the
% marginal covariance~$\Xi$ as a proxy for the true covariance
% function~$\Sigma_t(\cdot, \cdot)$.
% We found this to work well in simulations as~$\Sigma_t$ was relatively
% constant across~$t$; however, this imputation strategy may be
% inappropriate when the covariance structure changes significantly with
% time~$t$.

The imputation strategy generalizes the procedure introduced by~\cite{Petrovich2018} to the longitudinal functional data setting.  This is only possible due to the connection established with logistic regression by Lemma~\ref{lemma:logistic}. We could consider a multiple imputation estimate with~$M$ imputations; the likelihood estimate is given by~$\bar \theta = M^{-1} \sum_{m=1}^M \hat \theta^{(m)}$ where~$\theta^{(m)}$ is the estimate from the~$m$th imputated dataset; the covariance matrix can be estimated by
\[
M^{-1} \sum_{m=1}^M \hat \Xi^{(m)} (\theta) + \frac{1}{M-1}
\sum_{m=1}^M (\hat \theta^{(m)} - \bar \theta) (\hat \theta^{(m)} -
\bar \theta)
\]
where~$\hat \Xi^{(m)}$ is the estimator~\eqref{eq:fisher} for
the~$m$th imputated dataset.

\subsubsection{Bootstrap and multiple imputation under uncongenial }

The above procedure is

Positives: (1) our imputation model is ``approximately'' congenial; (2) our imputatoin model is simple and easy to use.
Negatives: (1) clearly not congenial under many events and/or high sampling rates. (2) difficult when we overlap due to over-sampling.

Congeniality matters because we want to ensure good coverage
Q: Can we derive a variance that provides coverage
We know our imputation models are uncongenial albeit simple to implement.  Impact should be small if events are rare as impute only in ranges where 1,0 labels give you everything.  To be careful, we investigate alternatives and assess impact.

Recommendation from~\cite{Bartlett2020}.

Simple version compared to the von Hippel's boot MI proposal.
\begin{itemize}
\item Bootstrap $B$ times
\item For $b = 1, \ldots, B$, impute $M$ times
\item Let~$\hat \theta_b = M^{-1} \sum_m \hat \theta_{b,m}$
\item Compute $\text{Var}_{BootMI} = (B-1)^{-1} \sum_{b=1}^B (\hat \theta_b - \hat \theta_{BM})^2$ where $\hat \theta_{BM} = B^{-1} \sum_{b=1}^B \hat \theta_b$.
\end{itemize}

@@

\begin{equation}
\label{eq:multilevel}
h_i \left( t \given  H_{i,t}^{NX} ; \theta, b_i \right) \approx
\exp \left( Z_{t}^\top \gamma + g_t \left( H_{i,t}^{N} \right)^{\prime} \alpha
  + \left[ M_{i,t}^\top + C_{i,t}^\top J_{\hat \psi, \phi}\right] (\beta + b_i) \right)
= \exp \left( W_{i,t}^\top \theta  + Z_{i,t}^\top b_i \right),
\end{equation}

@@

During a XX month period in 2019, 91 psychiatric inpatients admitted for suicidal risk to Franciscan Children's Hospital were enrolled in an observational study.  Study data were graciously provided by Evan Kleiman and his study team ({\tt https://kleimanlab.org}). Each study participant wore an Empatica E4~\cite{empaticae3}, a medical-grade wearable device that offers real-time physiological data.  One of the primary study goals was to assess the association between sensor-based physiological measurements and self-identified

@@

>>>>

In this section, we extend our model to the case of multiple functional regressors.  That is, suppose $L$ health processes, i.e., $\bfx_i = \{ \bfx_i (t) = (x_{i,1} (t), \ldots, x_{i,L} (t)) \}_{0 < s < \tau_i}$, for each participant is measured at a dense grid of time points. In the suicidal ideation case study, for example, accelerometer is measured at a rate of 32Hz while electrodermal activity (EDA) is measured at a rate of 4Hz.  A multivariate extension of our model~\eqref{eq:hazard} is given by
\begin{equation}
\label{eq:multihazardlinear}
h_i \left( t \given  H_{i,t}^{NX} ; \theta \right) =
h_0 (t; \gamma) \exp \left( g_t \left( H_{i,t}^{N} \right)^{\prime} \alpha
  + \int_{t-\Delta}^t x_{i,1} (s) \beta_1(s) ds + \cdots + \int_{t-\Delta}^t x_{i,L} (s) \beta_2(s) ds  \right)
\end{equation}
The approach given in Section~\ref{eq:logistication} extends naturally to the multivariate functional setting. For each functional regressor, we estimate the pooled sample covariance $\Sigma_{y,l}$ for $y \in \{0,1\}$ and $l=1,\ldots,L$ as in Section~\ref{sec:margcov}.  Let $\sum_{k=1}^\infty \hat \lambda^{(y)}_{k,l} \hat \psi^{(y)}_{k,l} (s) \hat \psi_{k,l}(t)$ be the spectral decomposition of $\hat \Sigma_{y,l}$.  Then $x_{i,j}(t)$ is approximated using a truncated Karhunen-Loeve
$$
\int_{t-\Delta}^t X_{l} (t,s) \beta(t) = \left[M_{l,t} + {\bf c}_l (t)^\top J_{\hat \psi_l^{(y)}, \phi} \right] {\bf b}_l
$$
Note that we express each coefficient function in the same spline basis; however, different basis expansions are used for each health process as defined by the marginal covariance calculation in Section~\ref{sec:margcov}.

@@

Here we present an extension of~\cite{Petrovich2018} to our setting by leveraging Lemma~\ref{lemma:logistic} and the marginal covariance estimation procedure to construct a multiple imputation procedure. Let~$\bfx_{i} (t)$ denote incomplete sensor data at time~$t$ (i.e., at times~$\{ s_{i,r} \}_{r=1}^{k_{it}}$ in~$[0,\Delta]$. Then
\begin{align}
\label{eq:impute}
\E [ X_i (s,t) \given Y_i(t) = y, \bfx_i (t) ]
  &= \mu_y (s,t) + {\bf a}_{i,t}^\top (s) {\bf B}_{i,t} (\bfx_i (t) -
    \mathbold{\mu}_i (t) ) \\
\text{Cov} \left( X_i (s,t), X_i (s^\prime, t) \given Y_i (t) = y,
  \bfx_i (t) \right)
  &= \Sigma_{t} (s, s^\prime) -
    {\bf a}_{i,t} (s)^\top {\bf B}_{i,t} {\bf a}_{i,t} (s^\prime)
\end{align}
where we have
\[
{\bf a}_{i,t} (s)^\top = \left( \begin{array}{c} \Sigma_t (s_{i,1}, s) \\ \vdots \\
                       \Sigma_t (s_{i,k_{it}}, s) \end{array} \right); \quad
{\bf B}_{i,t}^{-1} = \left(
  \begin{array}{c c c}
    \Sigma_t(s_{i,1}, s_{i,1}) & \Sigma_t(s_{i,1}, s_{i,2}) & \cdots \\
    \Sigma_t(s_{i,1}, s_{i,2}) & \ddots & \vdots \\
    \vdots & \cdots & \Sigma_t (s_{i,k_{it}}, s_{i,k_{it}})
  \end{array} \right),
\]
$\mathbold{\mu}_i (t) = \E [ \bfx_i (t) \given Y_i (t) = y] = \{
\mu_y (s_{j}, t) \}_{j=1}^r$ and~$\mu_y (s,t)$ is the mean of~$X(t,s)$
from group~$y$, and~$\Sigma_t ( s^\prime, s)$ is the covariance
between~$X(s^\prime,t)$ and $X(s,t)$ for $s^\prime, s \in [0, \Delta]$
and~$t \in \mathbb{R}_+$.
These covariances can be estimated using the methods considered by
\cite{ChenMuller2012}.
% Our imputation procedure instead uses the
% marginal covariance~$\Xi$ as a proxy for the true covariance
% function~$\Sigma_t(\cdot, \cdot)$.
% We found this to work well in simulations as~$\Sigma_t$ was relatively
% constant across~$t$; however, this imputation strategy may be
% inappropriate when the covariance structure changes significantly with
% time~$t$.

The imputation strategy generalizes the procedure introduced by~\cite{Petrovich2018} to the longitudinal functional data setting.  This is possible due to the connection established with logistic regression by Lemma~\ref{lemma:logistic}. We could then consider a multiple imputation estimate with~$M$ imputations; the likelihood estimate is given by~$\bar \theta = M^{-1} \sum_{m=1}^M \hat \theta^{(m)}$ where~$\theta^{(m)}$ is the estimate from the~$m$th imputated dataset; the covariance matrix can be estimated by
\[
M^{-1} \sum_{m=1}^M \hat \Xi^{(m)} (\theta) + \frac{1}{M-1}
\sum_{m=1}^M (\hat \theta^{(m)} - \bar \theta) (\hat \theta^{(m)} -
\bar \theta)
\]
where~$\hat \Xi^{(m)}$ is the estimator~\eqref{eq:fisher} for
the~$m$th imputated dataset.

\subsubsection{Multiple imputation under uncongeniality}

Multiple imputation yields valid frequentist inferences when the imputation and analysis procedure are congenial~\citep{Meng1994};  the above procedure was derived for function-on-scalar multiple imputation for binary outcomes, which ignored the joint nature of recurrent event analysis in the presence of high frequency sensor data. Consider, for example, two non-events sampled five minutes apart.  If there is missing data in the overlapping part of the temporal windows, then the above procedure imputes different values for each record.  Moreover, the imputatoin does not depend on whether one of the two were an event rather than non-event.

The main positive of the above procedure is that the above imputation framework is simple and approximately congenial when events are rare and the sampling rate is low as well as under the null of $\beta (s) \equiv 0$.  The main negative is that the model is clearly uncongenial under many events and/or high sampling rates. Again, congeniality ensures goo frequentist coverage properties.  A key question is whether we can use the above imputation methods within a general procedure to handle uncongeniality.

To address this, we use the recommendation from~\cite{Bartlett2020} and consider a method that first bootstraps a sample from the randomly sampled dataset and then apply multiple imputation to each bootstrap sample.  This general approach was originally proposed by~\cite{Shao1994} and~\cite{Little2002}. We suppose $B$ bootstraps and $M$ imputations per bootstrap; let $\hat \theta_{b,m}$ denote the estimator for the $m$th imputation of the $b$th bootstrap.  The point estimator is given by $(B)^{-1} \sum_{b=1}^B \hat \theta_{b}$ where $\hat \theta_b = M^{-1} \sum_{m=1}^M \hat \theta_{b,m}$. To construct the confidence interval, we require mean sum of squares with and between boostraps, i.e., $MSW = \frac{1}{B(M-1)} \sum_{b=1}^B \sum_{m=1}^M (\hat \theta_{b,m} - \hat \theta_b )^2$ and $MSB = \frac{1}{B-1} \sum_{b=1}^B M \cdot (\hat \theta_b - \hat \theta)^2$ respectively. Then the estimator of the variance of $\hat \theta$ is given by $\left( \frac{B+1}{BM} \right) MSB - MSW/M$.

The bootstrap followed by multiple imputation procedure has been studied extensively by~\cite{Bartlett2020} and is robust to uncongeniality.  The main disadvantage of this approach is its considerable computational intensity.  Recall likelihood calculations were computationally prohibitive by themselves, so combining with bootstrap and MI would further increase this large-scale computation.  The random subsampling framework thus simplifies handling of missing data via connections to function-on-scalar multiple imputation by~\cite{Petrovich2018} as well as to bootstrap to handle uncongeniality by~\cite{Bartlett2020}.

@@

In this section, we extend our model to the case of multilevel models with  functional regressors. Multilevel models are critical in mobile health where we expect high degree of individual variation.  In this case, we suppose that there is individual variation $b_i \sim N(0, \sigma_{b}^2)$ such that the multilevel extension of our model~\eqref{eq:hazard} is given by
\begin{equation}
\label{eq:multilevel}
h_i \left( t \given  H_{i,t}^{NX} ; \theta, b_i \right) \approx
\exp \left( Z_{t}^\top \gamma + g_t \left( H_{i,t}^{N} \right)^{\prime} \alpha
  + \left[ M_{i,t}^\top + C_{i,t}^\top J_{\hat \psi, \phi}\right] (\beta + b_i) \right)
= \exp \left( W_{i,t}^\top \theta  + Z_{i,t}^\top b_i \right),
\end{equation}
where $Z_Pi,t$

Ignoring the computational time of bootstrap sampling, the computational time of $200$ boostraps and $2$ imputations per boostrap leads to $35$ hours compared to $7$ hours.  Again, this highlights the substantial benefits of

@@

During a XX month period in 2019, 91 psychiatric inpatients admitted for suicidal risk to Franciscan Children's Hospital were enrolled in an observational study.  Study data were graciously provided by Evan Kleiman and his study team ({\tt https://kleimanlab.org}). Each study participant wore an Empatica E4~\cite{empaticae3}, a medical-grade wearable device that offers real-time physiological data.  One of the primary study goals was to assess the association between sensor-based physiological measurements and self-identified moments of suicidal distress.

@@

\section{Optimal weights}
